Improving \textsl{Gaia} parallax precision \\ with a data-driven model of stars

### 2017-06-14 paper/paper.tex
some of my corrections

We use stars crossmatched in \tgas\ and \tmass.
The match was done using a nearest-neighbor algorithm with a search radius of $4$ arcsec\footnote{\url{http://portal.nersc.gov/project/cosmo/temp/dstn/gaia/tgas-matched-2mass.fits.gz}}.
We also required that the stars lie within the
observing footprint of \psone\ to access the \cite{green15} 3D dust model.
We require that the photometry have real values, and nonzero, real errors, and
remove a small selection of \tmass\ stars that have zero color and zero $J$-band
magnitude.
The full data set is visualized in the \cmd\ in \figurename~\ref{fig:data}.
The left panel shows the point estimates of the color and absolute magnitude,
using the point estimate of the parallax from \gaia. The $1\sigma$ and
$2\sigma$ contours of the distribution are shown in grey.
The right panel shows a subset of the data with its associated error bars, calculated via subsampling.
The uncertainties in the colors are fairly well behaved, but the large
uncertainties in some absolute magnitudes are due to the large uncertainties in
the parallax measurements.

@@

\subsection{The empirical prior}
Using \xd\ and the MMLE method described in the methods section, we generate the
prior, shown in \figurename~\ref{fig:prior}.
The left panel shows 1.3 million samples from the prior distribution, and the
right panel shows $1\sigma$ contours for each component in the underlying
Gaussian mixture model.
Compared to the raw data in \figurename~\ref{fig:data}, the deconvolved
color--magnitude diagram for the \gaia\ + \tmass\ stars is tighter, as expected.
We'll come back to more features in the data below, especially those seen in the posterior distribution.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{prior.pdf}
\caption{The \cmd\ prior, modeled as a Gaussian mixture inferred by
  running XD on all of the data. The left panel shows a sampling of
  the prior, with the blue lines showing the $1\sigma$ and $2\sigma$
  contours. The right panel shows the 1-$\sigma$ contours of the
  individual components. The latter are Gaussian in the transformed
  magnitude space to make the XD inference tractable. Thus, they
  appear as slightly deformed ellipses in color--magnitude space.}
\label{fig:prior}
\end{figure}

@@

Figure~\ref{fig:posterior} shows the various probability distributions of our inference for a few randomly-chosen objects: the likelihood (black), prior (green), and posterior PDF (blue) on the true parallaxes.
For the majority of the \gaia\ stars, our prior slightly increases the precision of the posterior PDF compared with the \gaia\ variance. Our parallax posterior PDFs are more precise than the \tgas\ catalog entries by a median factor of $1.2$. Examples of these slight increases in precision are visualized in Examples 1 through 4 in Figure~\ref{fig:posterior} which show slightly narrower posterior PDFs compared to the likelihoods.
However, $14\%$ of our parallax posterior PDFs are more precise than the \tgas\ catalog by a factor of $>2$. This larger change in precision is visualized in Example 5, which shows a significantly narrower posterior PDF compared to the likelihood.
The nature of our prior is also visible: it is apparent that it is made of a mixture of components, and that its impact, and shape, strongly depend on the location of each star in \cmd\ space.

Figure~\ref{fig:posteriorCMD} shows the distribution of the expectation values of the posterior parallax PDFs projected back onto the \cmd\ space.
The left panel shows the distribution for all 1.3 million stars in our dataset, and the right panel shows a subsample of the dataset but includes the $1\sigma$ uncertainties on the posterior PDFs.
Compared with the raw data shown in Figure~\ref{fig:data}, it is clear the precision of the posterior PDFs is greater, especially for the red giant branch stars.

This change in precision is further illustrated in Figure~\ref{fig:delta}, where we show the fractional changes in the variance. The left panel shows the natural log of the fractional change in variance as a function of $(J-K_s)^C$ color, with the $1\sigma$ and $2\sigma$ contours over plotted.
The points are colored by the natural log of their fractional change in variance (the $y$ axis of the left panel) in both panels to help guide your eye for the right panel. The regions of \cmd\ space with the greatest improvements are shown in black, and the regions with the lowest improvements are shown in yellow.
The giants show obvious large improvements, as well as some main sequence stars.
Regions of large improvement are those with the largest photometric and parallax errors, which are the faintest observed objects.
This is a fairly standard result which we illustrated in the toy model above: the deconvolution of uncertainties is stronger in noisier regions of the data, leading to narrower features in the upper part of the \cmd, for example.
Stars with a color $(J-K_s)^C \sim 0.6$ show a decrease in precision due to the width of the prior there; the more vertical structure in the red giant branch of the \cmd.

Figure~\ref{fig:deltaCDF} shows the cumulative distribution function of the natural log of the fractional change in variance of the parallax posterior pdf relative to the \gaia\ catalog. This more directly shows what was alluded to in Figure~\ref{fig:posterior}. For the majority of the \gaia\ stars, our prior slightly increases the precision of the posterior PDF compared with the \gaia\ variance. Our parallax posterior PDFs are more precise than the \tgas\ catalog entries by a median factor of $1.2$, shown as the dotted line at $y=0.5$.
The precision of our posterior PDFs is greater than or equal to the precision of the \gaia\ catalogue for $87\%$ of the stars, shown as the dotted line at $x=0$,
and $14\%$ of our parallax posterior PDFs are more precise than the \tgas\ catalog by a factor of $>2$, which corresponds to a fraction decrease in variance by a factor of $>4$, shown as the dotted line at $y=0.14$.

@@

\subsection{Distances to M67}
As a test of the accuracy of these photometric parallax inferences, we inferred the parallaxes to stars in the open cluster M67.

M67 is at a Heliocentric distance of $800$--$900~{\rm pc}$ with an estimated age of $3$--$5~{\rm Gyr}$.
Previous work \citep{cbj15} has noted that distance estimates directly derived from the inverse of the observed (noisy) parallax are unreliable (significantly biased and uncertain) when the parallax signal-to-noise (SN) is below 5, $\varpi / \sigma_\varpi < 5$.

@@

We select sources from our catalog within a $1~{\rm deg}^2$ window centered on the position of M67; the cluster is visible as an over-density of stars in the sky positions alone.
To be clear, we do no prior selection on previously known members of M67, but include all stars observed by \gaia\ within this window on the sky.
Keep in mind, the (low) number of M67 members seen here is constrained by the (unknown) selection effects of the \tgas\ dataset.
With our converged dust values, and its associated prior, we calculate the posterior over $\varpi_{\true}$ for each star within this window on the sky around M67.
\figurename~\ref{fig:m67} shows the comparison of the likelihoods and posteriors.
The vertical lines bracket the previously measured parallaxes of, or distances to, the cluster.
The likelihoods of the observed parallaxes, shown in the top row, are broad and the cluster is not at all obvious.
The posterior PDFs, shown in the bottom row, are more sharply peaked showing an increase in precision.
The posterior PDFs also have modes that are more similar to the previously determined distance to the cluster (vertical lines) for far more stars.
This shows that our prior is not only making parallaxes more precise but also (possibly) increasing the accuracy.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{m67.pdf}
\caption{Proof of concept: parallax inference to the open cluster M67.
The upper left panel visualizes the \gaia\ + \tmass\ likelihoods of the data, shown in black and projected onto the \cmd, for all the stars within our $1~{\rm deg}^2$ window centered on the position of M67.
Our prior, visualized as a mixture of Gaussians, is shown in green.
The upper right panel shows the \gaia\ parallax likelihoods in black, with the ``true" distance to M67, determined by previous methods, bracketed by the red lines.
Similar to the upper left panel, the lower left panel shows the likelihoods of the data in black and our prior in green, as well as the posterior PDFs in blue.
The posterior PDFs are a mixture of Gaussians as well, each a product of the likelihood with a Gaussian in our prior.
The lower right panel shows the projection of the 2D posterior PDF onto parallax space.
The posterior PDFs are much more narrow and clustered than the \gaia\ likelihoods, showing that our prior generates posterior PDFs that are accurate and more precise than \gaia\ alone.}
\label{fig:m67}
\end{figure}

@@

There are other options for prior information to include when inferring distances to stars.
One option is the exponentially declining spatial density (EDSD) of stars \citep{astraatmadja16a}.
It is parameterized by a scale length, and has the nice property that it is smooth out to very large distances so that the posteriors are also very smooth.
It is a fairly weak prior.
\figurename~\ref{fig:comparePrior} shows a comparison of the EDSD prior with our \cmd\ prior.
Here the parallax posterior PDF expectation value and square root of the variance is converted into an absolute magnitude and visualized as the \cmd.
The left panel visualizes results using our \cmd\ prior, and the right panel visualizes results using the EDSD prior.
The variance for the \cmd\ prior is significantly smaller than the variance for the EDSD prior.
In general, it is a fairly broad prior so the median \gaia\ star has a posterior variance equal to the likelihood variance.
It has the largest effect on the most noisy and negative parallaxes, and has the nice property of bringing these noisy or negative measurements to reasonable parallaxes.

@@

  the \tgas--\tmass\ intersection, the model \emph{is} representative of

@@

  \tgas--\tmass\ intersection without any selection function restricts

@@

  \tgas--\tmass\ intersection.  It would be biasing to use this same

@@

  bluer than those included in the \tgas--\tmass\ intersection.

@@

  Those inhomogeneities are unlikely to affect the \cmd results; we effectively model the average distribution of the \tgas--\tmass\ intersection.

@@

  be easily optimized. We performed some experiments

@@

  conservative sense of cross-validation score. Choosing an optimized or

>>>>

We use stars crossmatched in \tgas\ and \tmass.
The match was done using a nearest-neighbor algorithm with a search radius of $4$ arcsec\footnote{\url{http://portal.nersc.gov/project/cosmo/temp/dstn/gaia/tgas-matched-2mass.fits.gz}}.
We also required that the stars lie within the
observing footprint of \psone\ to access the \cite{green15} 3D dust model.
We require that the photometry have real values, and nonzero, real errors, and
remove a small selection of \tmass\ stars that have zero color and zero $J$-band
magnitude.
The full data set is visualized in the \cmd\ in \figurename~\ref{fig:data}.
The left panel shows the point estimates of the color and absolute magnitude,
using the point estimate of the parallax from \gaia. The $1\sigma$ and
$2\sigma$ contours of the distribution are shown in grey.
The right panel shows a subset of the data with the associated error bar for each star.
The uncertainties in the colors are fairly well behaved, but the large
uncertainties in some absolute magnitudes are due to the large uncertainties in
the parallax measurements.

@@

\subsection{The empirical prior}
Using \xd\ and the MMLE method described in the methods section, we generate the
prior, shown in \figurename~\ref{fig:prior}.
The left panel shows 1.4 million samples from the prior distribution, and the
right panel shows $1\sigma$ contours for each component in the underlying
Gaussian mixture model.
Compared to the raw data in \figurename~\ref{fig:data}, the deconvolved
color--magnitude diagram for the \gaia\ + \tmass\ stars is tighter, as expected.
We'll come back to more features in the data below, especially those seen in the posterior distribution.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{prior.pdf}
\caption{The \cmd\ prior, modeled as a Gaussian mixture inferred by
  running XD on all of the data. The left panel shows a sampling of
  the prior, with the blue lines showing the $1\sigma$ and $2\sigma$
  contours. The right panel shows the 1-$\sigma$ contours of the
  individual components. The latter are Gaussian in the transformed
  magnitude space to make the XD inference tractable. Thus, they
  appear as slightly deformed ellipses in color--magnitude space.}
\label{fig:prior}
\end{figure}

@@

Figure~\ref{fig:posterior} shows the various probability distributions of our inference for a few randomly-chosen objects: the likelihood (black), prior (green), and posterior PDF (blue) on the true parallaxes.
For the majority of the \gaia\ stars, our prior slightly increases the precision of the posterior PDF compared with the \gaia\ likelihood. Our parallax posterior PDFs are more precise than the \tgas\ catalog entries by a median factor of $1.2$. Examples of these slight increases in precision are visualized in Examples 1 through 4 in Figure~\ref{fig:posterior} which show slightly narrower posterior PDFs compared to the likelihoods.
However, $14\%$ of our parallax posterior PDFs are more precise than the \tgas\ catalog by a factor of $>2$. This larger change in precision is visualized in Example 5, which shows a significantly narrower posterior PDF compared to the likelihood.
The nature of our prior is also visible: it is apparent that it is made of a mixture of components, and that its impact, and shape, strongly depend on the location of each star in \cmd\ space.

Figure~\ref{fig:posteriorCMD} shows the distribution of the expectation values of the posterior parallax PDFs projected back onto the \cmd\ space.
The left panel shows the distribution for all 1.4 million stars in our dataset, and the right panel shows a subsample of the dataset and includes the $1\sigma$ uncertainties on the posterior PDFs.
Compared with the raw data shown in Figure~\ref{fig:data}, it is clear the precision of the posterior PDFs is greater, especially for the red giant branch stars.

This change in precision is further illustrated in Figure~\ref{fig:delta}, where we show the fractional changes in the variance. The left panel shows the natural log of the fractional change in variance as a function of $(J-K_s)^C$ color, with the $1\sigma$ and $2\sigma$ contours of the distribution over plotted.
The points are colored by the natural log of their fractional change in variance (the $y$ axis of the left panel) in both panels to help guide your eye for the right panel. The regions of \cmd\ space with the greatest improvements are shown in black, and the regions with the lowest improvements are shown in yellow.
The giants show obvious large improvements, as well as some main sequence stars.
Regions of large improvement are those with the largest photometric and parallax errors, which are the faintest observed objects.
This is a fairly standard result which we illustrated in the toy model above: the deconvolution of uncertainties is stronger in noisier regions of the data, leading to narrower features in the upper part of the \cmd, for example.
Stars with a color $(J-K_s)^C \sim 0.6$ show a decrease in precision due to the width of the prior there; the more vertical structure in the red giant branch of the \cmd.

Figure~\ref{fig:deltaCDF} shows the cumulative distribution function of the natural log of the fractional change in variance of the parallax posterior pdf relative to the \gaia\ catalog (the y-axis of Figure~\ref{fig:delta}). This more directly shows what was alluded to in Figure~\ref{fig:posterior}. For the majority of the \gaia\ stars, our prior slightly increases the precision of the posterior PDF compared with the \gaia\ variance. Our parallax posterior PDFs are more precise than the \tgas\ catalog entries by a median factor of $1.2$, shown as the dotted line at $y=0.5$.
The precision of our posterior PDFs is greater than or equal to the precision of the \gaia\ catalogue for $87\%$ of the stars, shown as the dotted line at $x=0$,
and $14\%$ of our parallax posterior PDFs are more precise than the \tgas\ catalog by a factor of $>2$, which corresponds to a fraction decrease in variance by a factor of $>4$, shown as the dotted line at $y=0.14$.

LMA: describe Figure \ref{fig:deltaParallax}

@@

\subsection{Distances to M67}
As a test of the accuracy of these photometric parallax inferences, we inferred the parallaxes to stars in the open cluster M67.
M67 is at a Heliocentric distance of $800$--$900~{\rm pc}$ with an estimated age of $3$--$5~{\rm Gyr}$.
Previous work \citep{cbj15} has noted that distance estimates directly derived from the inverse of the observed (noisy) parallax are unreliable (significantly biased and uncertain) when the parallax signal-to-noise (SN) is below 5, $\varpi / \sigma_\varpi < 5$.

@@

We select sources from our catalog within a $1~{\rm deg}^2$ window centered on the position of M67; the cluster is visible as an over-density of stars in the sky positions alone.
To be clear, we do no prior selection on previously known members of M67, but include all stars observed by \gaia\ within this window on the sky.
Keep in mind, the (low) number of M67 members seen here is constrained by the (unknown) selection effects of the \tgas\ dataset.
With our converged dust values, and its associated prior, we calculate the posterior over $\varpi_{\true}$ for each star within this window on the sky around M67.
\figurename~\ref{fig:m67} shows the comparison of the likelihoods and posteriors.
The vertical red lines bracket the previously measured parallaxes of, or distances to, the cluster.
The likelihoods of the observed parallaxes, shown in the top row in black, are broad and the cluster is not at all obvious.
The posterior PDFs, shown in the bottom row in blue, are more sharply peaked showing an increase in precision.
The posterior PDFs also have modes that are more similar to the previously determined distance to the cluster (vertical lines) for far more stars.
This shows that our prior is not only making parallaxes more precise but also (possibly) increasing the accuracy.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{m67.pdf}
\caption{Proof of concept: parallax inference to the open cluster M67.
The upper left panel visualizes the \gaia\ + \tmass\ likelihoods of the data, shown in black and projected onto the \cmd, for all the stars within our $1~{\rm deg}^2$ window centered on the position of M67.
Our prior, visualized as a mixture of Gaussians, is shown in green.
The upper right panel shows the \gaia\ parallax likelihoods in black, with the ``true" distance to M67, determined by previous methods, bracketed by the red lines.
Similar to the upper left panel, the lower left panel shows the likelihoods of the data in black and our prior in green, as well as the posterior PDFs in blue.
The posterior PDFs are a mixture of Gaussians as well, each a product of the likelihood with a Gaussian in our prior.
The lower right panel shows the projection of the 2D posterior PDF onto parallax space.
The posterior PDFs are much more narrow and clustered than the \gaia\ likelihoods, showing that our prior generates posterior PDFs that are accurate and more precise than \gaia\ alone.}
\label{fig:m67}
\end{figure}

@@

There are other options for prior information to include when inferring distances to stars.
One option is the exponentially declining spatial density (EDSD) of stars \citep{astraatmadja16a}.
It is parameterized by a scale length, and has the nice property that it is smooth out to very large distances so that the posteriors are also very smooth.
It is a fairly weak prior.
\figurename~\ref{fig:comparePrior} shows a comparison of the EDSD prior with our \cmd\ prior.
Here the parallax posterior PDF expectation value, and $1\sigma$ uncertainty, for a subset of stars in our dataset is converted into an absolute magnitude and visualized as the \cmd.
The left panel visualizes results using the EDSD prior, and the right panel visualizes results using our \cmd\ prior.
The variance for the \cmd\ prior is significantly smaller than the variance for the EDSD prior.
In general, EDSD is a fairly broad prior so the median \gaia\ star has a posterior variance equal to the likelihood variance.
EDSD has the largest effect on the most noisy and negative parallaxes, and has the nice property of bringing these noisy or negative measurements to reasonable parallaxes.

@@

  the \tgas--\tmass--\psone\ intersection, the model \emph{is} representative of

@@

  \tgas--\tmass--\psone\ intersection without any selection function restricts

@@

  \tgas--\tmass--\psone\ intersection.  It would be biasing to use this same

@@

  bluer than those included in the \tgas--\tmass--\psone\ intersection.

@@

  Those inhomogeneities are unlikely to affect the \cmd\ results; we effectively model the average distribution of the \tgas--\tmass--\psone\ intersection.

@@

  easily optimized. We performed some experiments

@@

  conservative sense of a cross-validation score. Choosing an optimized or
